\begin{center} \bf \sffamily \LARGE Prisoner's dilemma with stochastic payoffs\\[-1cm] \end{center}

### 2023-03-06 paper/main.tex
finish reading paper

Similar to all other theoretical models, finite stochastic processes rest on a
set of assumptions. Namely, in the game phase it is assumed that players use
strategies with finite memory~\cite{Nowak1992tit, Baek2016}. For example, in a
repeated game between two individuals the actions of the players in each turn
are often determined by the action of the co-players in the previous turn~\cite{Nowak1992tit}. This
assumption is common as it allows for an explicit calculation of the players'
payoffs~\cite{sigmund2010calculus}. In addition, it is
assumed that individuals play many times and with all other players before
reproduction takes place. So that in the updating phase the updating payoffs,
or fitness, of individuals is given by the average payoffs of an
infinitely repeated game against every member of the population.

The above two assumptions create a curious inconsistency; a player with a finite
memory in the game stage can recall infinite information regarding their
interactions and each interaction's outcomes in the update stage. Previous work
has explored the effects of constraining the interactions an individual has.
Either by letting individuals have a stochastic number of
interactions~\cite{roca:PhysicalReview:2006, sanchez:JTB:2005,
Traulsen:JTB:2007} or by only allowing for
a single interaction~\cite{Woelfing:JTB:2009}. However, no previous work
considers that not only the interactions are limited but also the information
regarding each interaction. To this end, we propose a framework in which
individuals, similar to the decisions at each turn, estimate their fitness based
on a minimum of information.

@@

In order to account for the effect of the updating payoffs we simulate the
evolutionary process and record which strategies the players adopt over time
based on perfect and limited payoff memory. We performed two independent runs
for each approach varying the value of benefit \(b\).
Figure~\ref{fig:expected_and_stochastic_for_donation} depicts the evolving
conditional cooperation probabilities $p$ and $q$ (since the discount
factor~$\delta$ is comparably high we do not report the opening move \(y\)).

The figure suggests that when updating is based on perfect payoff memory players
tend to be more generous and more cooperative. The $q$-values of the resident
strategies are on average higher in this case, thus, players will occasionally
forgive a defection more often if their fitness depends on interacting with
every member of the population. This effect becomes more notable as the value of
the benefit increases. In the case of limited memory the resident strategies are
less forgiving and their forgiveness regardless of the values of benefit. In
analysis carried out in the Supplementary Information Section 3 we show that a
resident population of conditional cooperators with \(q > \frac{1}{2}\) can not
avoid invasion by a defecting strategy. This holds independently of the cost and
benefit values.

Higher $q$ values lead to a more cooperative population. We calculate the
average cooperation rate for each simulation which is the average cooperation
rate within the resident population. In the case of the perfect payoff memory 
the average cooperation rate is strictly higher than that of the last round
payoffs. The difference between the cooperation rate is the highest at $b=10$.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.9\textwidth]{static/donation_expected_last_round_summary_results.pdf}
    \caption{{\bf Evolutionary dynamics under perfect and limited payoff memory.}
    ({\bf Schematic illustrations}) On the left panels we show schematic
    illustrations of the perfect memory and the limited memory cases. The shaded
    background denotes the game phase information that an individual considers
    when updating strategies. In the case of perfect payoff memory the entire region is
    shaded and in the case of limited payoff memory only one turn with a single member
    of the population. ({\bf Simulations}) We have run four simulations of the
    pairwise comparison process for $T\!=\!10^7$ time steps. For each time step
    of the process
    we record the current resident population ($y,p,q$). Since
    simulations are run for a relatively high continuation probability of
    $\delta\!=\!0.999$, we do not report the players' initial cooperation
    probability $y$. The plots show how often the resident population chooses
    each combination ($p,q$) of conditional cooperation probabilities in the
    subsequent rounds. We also report the evolved cooperation rate which is
    calculated as the average cooperation rate within the resident population.
    ({\bf Perfect Memory}) In the case of low benefit the resident population
    either consists of defectors (with $p\!\approx\!q\!\approx\!0$) or of
    conditional cooperators. Conditional cooperators, or otherwise known as
    generous tit for tat, are a set of strategies that always cooperate
    following a cooperation ($p\!\approx\!1\!$) and cooperate with a probability
    $q$ given that the co-player has defected. $q$ denotes the generosity of a
    player. The resident population applies a conditional cooperator strategy
    for which $q\!\le\!1\!-\!c/b\!=\!0.67$. 
    In the case of high benefit the population mainly consists of conditional
    cooperators of the form ($p\!\approx\!1\!, q\!\le\!1\!-\!1/10\!=\!0.9$). In
    the Supplementary Information Section 2 we show that a conditional
    cooperator needs to be of the form ($p\!\approx\!1\!, q\!\le\!1\!-\!c/b$) to
    not be invaded by defecting strategies. A higher generosity in the
    population results in a higher average cooperation rate. The average
    cooperation rate increases from 52\% for \(b=3\) to 98\% for \(b=10\). ({\bf
    Limited Memory}) When players update their strategies based on their
    realized payoffs in the last round, there are two different predominant
    behaviors regardless of the benefit value. The resident population either
    consists of defectors (with $p\!\approx\!q\!\approx\!0$) or of conditional
    cooperators. The maximum level of $q$, consistent with stable cooperation, is
    somewhat smaller compared to the perfect memory setting.
    Namely, in the Supplementary Information Section 3 we show that
    regardless of the value of benefit a conditional cooperator need to be of
    the form $q\!<\!\frac{1}{2}$ to not be invaded by defectors.
    The evolved cooperation
    rate only slightly increases from 37\% ($b=3$) to 51\% ($b=10$). Parameters: $N\!=\!100$,
    $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.999$.}
    \label{fig:expected_and_stochastic_for_donation}
\end{figure}

We further explore the effects of the benefit and the strength of
selection on the generosity \(q\) and on the cooperation rate
(Figure~\ref{fig:cooperation_rate_over_benefit_and_beta}).
Figure~\ref{fig:cooperation_rate_over_benefit_and_beta}\textbf{A} suggests that
perfect memory always yields a higher cooperation rate. In this case the
cooperation rate increases as the benefit increases. In comparison for limited
payoff memory the cooperation rate remains unchanged at approximately 50\% once
\(b=5\).  From Figure~\ref{fig:cooperation_rate_over_benefit_and_beta}\textbf{B}
we observe that for weak selection, \(\beta < 1\), the two methods yield similar
results, however, as \(\beta\) increases there is variation in the evolving
populations. In the case of expected payoffs the resident populations become
more cooperative, whereas in the case of limited payoff memory, the resident
populations become more defective.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.75\textwidth]{static/cooperation_rate_over_b_and_beta.pdf}
    \caption{{\bf The evolution of cooperation and generosity for different
    values of benefit (A) and strength of selection (B).} We report the average
    cooperation and the average reciprocity. The average cooperation rate is the
    average cooperation rate within the resident population. For the average
    reciprocity we select the residents that have a $p \approx 1$ and we take
    the average of their cooperation probability $q$. ({\bf A}) We vary the
    benefit of cooperation $b$. In all cases, perfect memory updating payoffs
    appear to overestimate the average cooperation rate the population achieves.
    As expected in the case of limited memory the average generosity over the
    different values of benefit remains the same ($q \approx 0.5$), and as a
    result so does the average cooperation. ({\bf B}) We vary the selection
    strength $\beta$. For weak selection, \(\beta < 1\), the two methods yield
    similar results. However, as \(\beta\) increases in the case of limited
    memory payoffs the resident populations become more defective. Unless
    explicitly varied, the parameters of the simulation are $N\!=\!100$,
    $b\!=\!3$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.99$. Simulations are run
    for $T\!=\!5\times 10^7$ time steps for each parameter
    combination.}\label{fig:cooperation_rate_over_benefit_and_beta}
\end{figure}

The framework of limited
payoff memory can be easily extended to allow individuals to observe more
rounds, to interact with more members or both (see Supplementary Information
Sections 4-6). To further understand the effect of limited payoff memory we
consider the cases of updating based on the last round with two members of the
population, the last two rounds with another member of the population, and the
last two rounds with two members of the population. We perform numerical
simulations for each of the fitness methods. In
Figure~\ref{fig:cooperation_rate_all_updating_payoffs} we report the cooperation
rates for low and high values of benefit.

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=\textwidth]{static/more_memory_summary_results.pdf}
  \caption{{\bf Average cooperation rates for different updating payoffs.}
  From left to right, we present result on the following updating payoffs cases;
  (a) the expected payoffs (perfect memory), (b) the last round payoff from one
  interaction (limited memory), (c) the last round payoff from two interactions,
  (d) the last two rounds payoffs from one interaction, (e) the last two rounds
  payoffs from two interactions. For the updating payoffs (c) and (d) we have
  carried out an analysis which has shown that conditional cooperators should
  adopt a \(q\) value smaller than $\frac{\delta - 1 +
  \frac{\sqrt{2}}{2}}{\delta}$ and $\frac{\delta + \sqrt{\delta^{2} + 1} - 1}{2 \delta}$
  respectively. Note that as \(\delta
  \rightarrow 1\) both right hand sides tend to \(\frac{\sqrt{2}}{2}\).
  Regardless the cooperation rate between for case (c) is slightly higher.
  In case (d) the cooperation rate is the second highest hinting that as we
  allow for more information the closer we move to the perfect payoff memory.
  Unless explicitly varied, the parameters of the simulation are
  $N\!=\!100$, $b\!=\!3$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.99$.
  Simulations are run for $T\!=\!5\times 10^7$ time steps for each parameter
  combination.}\label{fig:cooperation_rate_all_updating_payoffs}
\end{figure}

@@

Cooperation can be seen at odd, why is it that we choose to help others at a
personal cost? In spite of all the selfish genes', animal and human communities
show signs of altruism and cooperation~\cite{milinski1987tit, kerr2002local,
carter2020development}. Evolutionary game theoretical models have helped us
shape our understanding of the evolution of cooperation. In fact, the evolution
of cooperation constitutes such a major focus of the field that evolutionary
game theory seems to be reduced to the evolution of
cooperation~\cite{Traulsen:PhilTrans:2022}. Previous evolutionary models often
feature a curious inconsistency. When modeling how individuals make decisions in
each round, these models assume that players only remember the last round.
However, when modeling how individuals update their strategies over time,
individuals are assumed to have perfect memory.

Here, we have explored how robust cooperation is as models deviate from the
perfect memory assumption. We considered a special case of the prisoner's
dilemma, the donation game. We showed
that when the last round payoff is used instead of the expected payoffs,
cooperation struggles to evolve  when only a minimum of social information is
used at the updating stage. The resident strategies were both less cooperative
and less generous. This effect was only intensified we increase the benefit and
the strength of selection independently. The results showed that as each
parameter was being increased the difference between the cooperation rates were
widening. This indicates that cooperative players benefit from being able to
interact with everyone in the population.

We extended our approach and presented results on different limited memory
payoffs. The analysis showed cooperation can even evolve when we allow
individual to consider more information.

>>>>

Like all theoretical models, finite stochastic processes rest on a
set of assumptions. Namely, in the game phase it is assumed that players use
strategies with finite memory~\cite{Nowak1992tit, Baek2016}. For example, in a
repeated game between two individuals the actions of the players in each turn
are often determined by the action of the co-players in the previous turn~\cite{Nowak1992tit}. This
assumption is common as it allows for an explicit calculation of the players'
payoffs~\cite{sigmund2010calculus}. In addition, it is
assumed that individuals play many times and with all other players before
reproduction takes place. So that in the updating phase the updating payoffs,
or fitness, of individuals is given by the average payoffs of an
infinitely repeated game against every member of the population.

These two assumptions create a curious inconsistency; a player with a finite
memory (in the game stage) can recall infinite information regarding their
interactions and each interaction's outcomes in the update stage. Prior research
has examined the effects of constraining the interactions an individual has.
Either by letting individuals have a stochastic number of
interactions~\cite{roca:PhysicalReview:2006, sanchez:JTB:2005,
Traulsen:JTB:2007} or by only allowing for a single
interaction~\cite{Woelfing:JTB:2009}. However, these studies have not accounted
for the use of strategies with finite memory. Moreover, no previous research has
explored the effect of restricting not only the number of interactions, but also
the amount of information available about each interaction. To this end, we
propose a framework in which individuals, similar to the decisions at each turn,
estimate their fitness based on a minimum of information.

@@

To assess the impact of updating payoffs, we simulate the evolutionary process,
recording the strategies adopted by players over time based on perfect and
limited payoff memory. We performed two separate runs for each approach varying
the value of benefit \(b\).
Figure~\ref{fig:expected_and_stochastic_for_donation} depicts the evolving
conditional cooperation probabilities $p$ and $q$ (note that we omit the opening
move \(y\), as the discount factor \(\delta\) is relatively high). The figure
suggests that when updating is based on perfect payoff memory players tend to be
more generous and more cooperative.

Specifically, we observe that the resident population comprises either defectors
or conditional cooperators \((1, q)\). The generosity level \(q\) adopted by the
resident population depends on whether a defecting strategy can invade.
Supplementary Information Section 2 and 3 show that in the perfect payoff memory
framework, conditional cooperators of the form \((1, q < \frac{c}{b})\) can
arise, however, in the case of limited payoff memory, only conditional
cooperators of \((1, q < \frac{1}{2})\) can avoid invasion. 
The $q$-values of the resident strategies are generally higher in classical
case, indicating that players are more likely to forgive a defection if their fitness
depends on interacting with every member of the population. This effect becomes
more pronounced as the benefit value increases, as the perfect memory condition
on the left-hand side increases, while in the limited memory case, it remains at
\(q \approx \frac{1}{2}\).

Higher $q$ values lead to a more cooperative population. We compute the average
cooperation rate for each simulation, which is the average cooperation rate
within the resident population. In the case of perfect payoff memory, the
average cooperation rate is consistently higher than that of the last round
payoffs.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=\textwidth]{static/donation_expected_last_round_summary_results.pdf}
    \caption{{\bf Evolutionary dynamics under perfect and limited payoff memory.}
    ({\bf Schematic illustrations}) On the left panels we show schematic
    illustrations of the perfect memory and the limited memory cases. The shaded
    background denotes the game phase information that an individual considers
    when updating strategies. In the case of perfect payoff memory the entire region is
    shaded and in the case of limited payoff memory only one turn with a single member
    of the population. ({\bf Simulations}) We have run four simulations of the
    pairwise comparison process for $T\!=\!10^7$ time steps. For each time step
    of the process
    we record the current resident population ($y,p,q$). Since
    simulations are run for a relatively high continuation probability of
    $\delta\!=\!0.999$, we do not report the players' initial cooperation
    probability $y$. The plots show how often the resident population chooses
    each combination ($p,q$) of conditional cooperation probabilities in the
    subsequent rounds. We also report the evolved cooperation rate which is
    calculated as the average cooperation rate within the resident population.
    ({\bf Perfect Memory}) In the case of low benefit the resident population
    either consists of defectors (with $p\!\approx\!q\!\approx\!0$) or of
    conditional cooperators. Conditional cooperators, or otherwise known as
    generous tit for tat, are a set of strategies that always cooperate
    following a cooperation ($p\!\approx\!1\!$) and cooperate with a probability
    $q$ given that the co-player has defected. $q$ denotes the generosity of a
    player. The resident population applies a conditional cooperator strategy
    for which $q\!\le\!1\!-\!c/b\!=\!0.67$. 
    In the case of high benefit the population mainly consists of conditional
    cooperators of the form ($p\!\approx\!1\!, q\!\le\!1\!-\!1/10\!=\!0.9$). In
    the Supplementary Information Section 2 we show that a conditional
    cooperator needs to be of the form ($p\!\approx\!1\!, q\!\le\!1\!-\!c/b$) to
    not be invaded by defecting strategies. A higher generosity in the
    population results in a higher average cooperation rate. The average
    cooperation rate increases from 52\% for \(b=3\) to 98\% for \(b=10\). ({\bf
    Limited Memory}) When players update their strategies based on their
    realized payoffs in the last round, there are two different predominant
    behaviors regardless of the benefit value. The resident population either
    consists of defectors (with $p\!\approx\!q\!\approx\!0$) or of conditional
    cooperators. The maximum level of $q$, consistent with stable cooperation, is
    somewhat smaller compared to the perfect memory setting.
    Namely, in the Supplementary Information Section 3 we show that
    regardless of the value of benefit a conditional cooperator need to be of
    the form $q\!<\!\frac{1}{2}$ to not be invaded by defectors.
    The evolved cooperation
    rate only slightly increases from 37\% ($b=3$) to 51\% ($b=10$). Parameters: $N\!=\!100$,
    $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.999$.}
    \label{fig:expected_and_stochastic_for_donation}
\end{figure}

We further investigate the impact of benefit and selection strength on
generosity $q$ and the cooperation rate, as shown in Figure~\ref{fig:cooperation_rate_over_benefit_and_beta}.
According to Figure~\ref{fig:cooperation_rate_over_benefit_and_beta}\textbf{A},
perfect memory consistently results in a higher cooperation rate, which
increases with increasing benefit. On the other hand, the cooperation rate
remains approximately 50\% for limited payoff memory once \(b = 5\).
From Figure~\ref{fig:cooperation_rate_over_benefit_and_beta}\textbf{B}
we observe that for weak selection, \(\beta < 1\), the two methods yield similar
results, however, as \(\beta\) increases there is variation in the evolving
populations. In the case of expected payoffs the resident populations become
more cooperative, whereas in the case of limited payoff memory, the resident
populations become more defective.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.75\textwidth]{static/cooperation_rate_over_b_and_beta.pdf}
    \caption{{\bf The evolution of cooperation and generosity for different
    values of benefit (A) and strength of selection (B).} We report the average
    cooperation and the average reciprocity. The average cooperation rate is the
    average cooperation rate within the resident population. For the average
    reciprocity we select the residents that have a $p \approx 1$ and we take
    the average of their cooperation probability $q$. ({\bf A}) We vary the
    benefit of cooperation $b$. In all cases, perfect memory updating payoffs
    appear to overestimate the average cooperation rate the population achieves.
    As expected in the case of limited memory the average generosity over the
    different values of benefit remains the same ($q \approx 0.5$), and as a
    result so does the average cooperation. ({\bf B}) We vary the selection
    strength $\beta$. For weak selection, \(\beta < 1\), the two methods yield
    similar results. However, as \(\beta\) increases in the case of limited
    memory payoffs the resident populations become more defective. Note that
    in the case of perfect payoff memory we see an increase in the cooperation
    rate even though the generosity remains stable. That is because the
    generosity does remain the same, however, now cooperative strategies remain
    fixed as the resident strategy for longer.
    Unless
    explicitly varied, the parameters of the simulation are $N\!=\!100$,
    $b\!=\!3$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.99$. Simulations are run
    for $T\!=\!5\times 10^7$ time steps for each parameter
    combination.}\label{fig:cooperation_rate_over_benefit_and_beta}
\end{figure}

The limited payoff memory framework can be expanded by enabling individuals to
observe a greater number of rounds, interact with a larger number of members, or
both (refer to Supplementary Information Sections 4-6). To gain further insight
into the impact of limited payoff memory, we explore the scenarios of updating
based on the last round with two members of the population, the last two rounds
with another member of the population, and the last two rounds with two members
of the population. To analyze the effects of this framework, we conduct
numerical simulations using various fitness methods. The cooperation rates for
low and high benefits are presented in Figure~\ref{fig:cooperation_rate_all_updating_payoffs}.

We observe a significant rise in the cooperation rate with the introduction of
slightly more information. Specifically, in scenarios involving two rounds or
two interactions, the cooperation rates are almost identical. For a large
population and a high continuation probability ($\delta$), the conditional
cooperators that are adopted by the resident population in these scenarios take
on the same form \((1, q < \frac{\sqrt{2}}{2})\). When considering both sets
of information (i.e., two rounds and two interactions), cooperation rates
experience a greater increase, yet still remain lower compared to the classical
scenario.

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=\textwidth]{static/more_memory_summary_results.pdf}
  \caption{{\bf Average cooperation rates for different updating payoffs.}
  From left to right, we present result on the following updating payoffs cases;
  (a) the expected payoffs (perfect memory), (b) the last round payoff from one
  interaction (limited memory), (c) the last round payoff from two interactions,
  (d) the last two rounds payoffs from one interaction, (e) the last two rounds
  payoffs from two interactions. For the updating payoffs (c) and (d) we have
  carried out an analysis which has shown that conditional cooperators should
  adopt a \(q\) value smaller than $\frac{\delta - 1 +
  \frac{\sqrt{2}}{2}}{\delta}$ and $\frac{\delta + \sqrt{\delta^{2} + 1} - 1}{2 \delta}$
  respectively. Note that as \(\delta
  \rightarrow 1\) both right hand sides tend to \(\frac{\sqrt{2}}{2}\).
  Regardless the cooperation rate between for case (c) is slightly higher.
  In case (d) the cooperation rate is the second highest hinting that as we
  allow for more information the closer we move to the perfect payoff memory.
  We performed four pairwise non parametric tests (Mann-Whitney U) to compare
  the cooperation distributions of the residents in case (b) to
  cases (a), (c), (d), (e). In all four tests we reject the null hypothesis
  with \(\alpha=0.05\) and \(p \approx 0\). Thus, there is significant
  difference between the cooperation rates.
  Unless explicitly varied, the parameters of the simulation are
  $N\!=\!100$, $b\!=\!3$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.99$.
  Simulations are run for $T\!=\!5\times 10^7$ time steps for each parameter
  combination.}\label{fig:cooperation_rate_all_updating_payoffs}
\end{figure}

@@

Cooperation can be seen as odd, why is it that we choose to help others at a
personal cost? In spite of all the selfish genes', animal and human communities
show signs of altruism and cooperation~\cite{milinski1987tit, kerr2002local,
carter2020development}. Evolutionary game theoretical models have helped us
shape our understanding of the evolution of cooperation. In fact, the evolution
of cooperation constitutes such a major focus of the field that evolutionary
game theory seems to be reduced to the evolution of
cooperation~\cite{Traulsen:PhilTrans:2022}.

Evolutionary models in the past often feature a curious inconsistency. While
these models depict how individuals make decisions in each round by assuming
that they only retain memory of the previous round, they also assume that
individuals possess a perfect memory when it comes to updating their strategies
over time. To be precise, individuals are assumed to remember all of their past
interactions and each interaction's outcome when updating strategies.

Here, we investigate the robustness of cooperation as models deviate
from the assumption of perfect memory. While prior research has investigated the
impact of constraining individuals' interactions, we take into account the
limitation of not only interactions but also the information available for each
outcome. Additionally, prior studies have only allowed for the adoption of
simple strategies such as always cooperating or always defecting. In contrast,
we enable the use of more intricate strategies where players can utilize the
previous play of their co-player to make decisions.

In our framework, players update their strategies based on a combination of
interactions and outcomes. The initial scenario we examined involved using one
piece of information: the last round of one interaction. The outcomes suggest
that cooperation faces difficulties in developing when the updating stage
utilizes minimal social information. This effect is compounded as the benefit
and strength of selection are independently increased. The findings indicate
that cooperative players benefit from the ability to engage with all members of
the population.

Furthermore, we investigated scenarios where the final two rounds, or the last
two instances of interaction, were taken into account. We observed a
statistically significant rise in the frequency of cooperative behavior. For a
sizable population and a high likelihood of continued interactions, the two
cases yield the same result as the overall payoff is influenced by two possible
outcomes. Notably, the scenario involving two rounds and two interactions
yielded the highest cooperation rate among all the novel methodologies that we
tested.
% A nice sentence missing here.
